---
title: "Humpback whale distributions and insight into preferred oceanographic conditions in the Caribbean using passive acoustic monitoring (PAM) and an autonomous underwater vehicle (AUV)"
author: "Keiley Gregory"
date: today
format:
  html:
    css: style.css
    theme: flatly
    highlight-style: tango
execute:
  working-directory: none
bibliography: references.bib
---

***COMMAND F \* FOR FIXES***

Humpback whale (*Megaptera novaeangliae*)

short background on:

how audio recorded by DMON2 and glider mission path

acoustic analysis in LFDCS and manual verification

raster calculations in arc pro

All quantitative oceanographic variables, including those measured directly by sensors on the glider (water temperature, salinity, dissolved oxygen concentration) and those calculated in ArcGIS Pro (bathymetric depth, bathymetric slope, distance to nearest shoreline), ...

future research - study will use data from more missions in different areas, compare findings between missions, overall patterns, blah blah

# Humpback Whale Distributions

short overview of findings from map (where detections clustered, geomorph type, etc.)

**Fig. 1.** \*WRITE CAPTION

\*link to interactive map

# Oceanographic Conditions

overview of findings for preferred oceanographic conditions

```{r, include=FALSE}
library(tidyverse)
library(ggridges) # for ridge plots
library(patchwork) # optional, for combining plots
library(knitr)
library(kableExtra) # optional but makes it prettier
```

```{r, include=FALSE}
# Import humpback detection data with GIS calculations
hbk <- read_csv("data/GISCALC_HUWH_AroundSTX2025.csv")

# Import background mission data
bkg <- read_csv("data/erddap_UVI2_Around_STX_20250306.csv")
```

```{r, include=FALSE}
# Rename relevant columns in BKG df so variable names match detections (HBK) df
bkg <- bkg %>% rename(
  "Water_temperature_C" = "water_temperature",
  "Salinity_PSU" = "salinity",
  "Oxygen_concentration_µmol_L" = "oxygen_concentration"
  )
```

```{r, include=FALSE}
# Vector of environmental variables to summarize for table (*EXCLUDE glider depth because not really an environmental variable)
env_vars <- c(
  "Water_temperature_C",
  "Salinity_PSU",
  "Oxygen_concentration_µmol_L",
  "RASTERDEPTH_m",
  "RASTERSLOPE_deg",
  "NEAR_DIST"
  )
```

```{r, include=FALSE}
# Create DETECTIONS summary table dataframe
env_summary <- hbk %>%
  select(all_of(env_vars)) %>%
  pivot_longer(
    cols = everything(),
    names_to  = "Variable",
    values_to = "Value"
  ) %>%
  group_by(Variable) %>%
  summarise(
    n     = sum(!is.na(Value)),
    Mean  = mean(Value, na.rm = TRUE),
    SEM   = sd(Value, na.rm = TRUE) / sqrt(n),
    SD    = sd(Value, na.rm = TRUE),
    Min   = min(Value, na.rm = TRUE),
    Max   = max(Value, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  mutate(
    Variable = case_when(
      Variable == "Water_temperature_C" ~ "Water Temperature (°C)",
      Variable == "Salinity_PSU" ~ "Salinity (PSU)",
      Variable == "Oxygen_concentration_µmol_L" ~ "Dissolved Oxygen Concentration (µmol/L)",
      Variable == "RASTERDEPTH_m" ~ "Bathymetric Depth (m)",
      Variable == "RASTERSLOPE_deg" ~ "Bathymetric Slope (°)",
      Variable == "NEAR_DIST" ~ "Distance to Nearest Shoreline (m)",
      TRUE ~ Variable
    )
  )

# Reorder rows for table
env_summary <- env_summary %>%
  mutate(
    Variable = factor(Variable,
    levels = c(
        "Water Temperature (°C)",
        "Salinity (PSU)",
        "Dissolved Oxygen Concentration (µmol/L)",
        "Bathymetric Depth (m)",
        "Bathymetric Slope (°)",
        "Distance to Nearest Shoreline (m)"
      )
    )
  ) %>%
  arrange(Variable)
```

```{r, include=FALSE}
# Create BACKGROUND summary table dataframe
bkg_summary <- bkg %>%
  select("Water_temperature_C", "Salinity_PSU", "Oxygen_concentration_µmol_L") %>%
  pivot_longer(
    cols = everything(),
    names_to  = "Variable",
    values_to = "Value"
    ) %>%
  group_by(Variable) %>%
  summarise(
    n     = sum(!is.na(Value)),
    Mean  = mean(Value, na.rm = TRUE),
    SEM   = sd(Value, na.rm = TRUE) / sqrt(n),
    SD    = sd(Value, na.rm = TRUE),
    Min   = min(Value, na.rm = TRUE),
    Max   = max(Value, na.rm = TRUE),
    .groups = "drop"
    ) %>%
  mutate(
    Variable = case_when(
      Variable == "Water_temperature_C" ~ "Water Temperature (°C)",
      Variable == "Salinity_PSU" ~ "Salinity (PSU)",
      Variable == "Oxygen_concentration_µmol_L" ~ "Dissolved Oxygen Concentration (µmol/L)",
      TRUE ~ Variable
      )
    )

# Reorder rows for table
bkg_summary <- bkg_summary %>%
  mutate(
    Variable = factor(Variable,
    levels = c(
      "Water Temperature (°C)",
      "Salinity (PSU)",
      "Dissolved Oxygen Concentration (µmol/L)"
      )
    )
  ) %>%
  arrange(Variable)
```

```{r, include=FALSE}
# Join env and background summaries by Variable
env_summary_diff <- env_summary %>%
  left_join(bkg_summary %>% 
              select(Variable, BKG_Mean = Mean), by = "Variable") %>%
  mutate(`Mean Difference from Background` = 
           sprintf("%+.5f", Mean - BKG_Mean))
```

```{r, include=FALSE}
# Reorder columns and remove "n" and "BKG_Mean" columns (*do NOT need to add '%>% select(- "n", - "BKG_Mean")' because Mean_bkg not included in first select code line so already removed from df)
env_summary_table <- env_summary_diff %>%
  select("Variable", "Mean", "SEM", "SD", "Min", "Max", "Mean Difference from Background")
```

**Table 1.** \*WRITE CAPTION

```{r, echo=FALSE}
env_summary_table |>
  kable(format = "html", digits = 2, align = "l") |>
  kable_styling(full_width = FALSE, bootstrap_options = c("striped", "hover"))
```

```{r, include=FALSE}
# Create custom color scheme for oceanographic variables
custom_colors <- c(
  Water_temperature_C = "#7600d1",
  Salinity_PSU = "#1002bd",
  Oxygen_concentration_µmol_L = "#908bc9",
  RASTERDEPTH_m = "#a45cad",
  RASTERSLOPE_deg = "#e632ad",
  NEAR_DIST = "#ffa100"
  )
```

```{r, include=FALSE}
# Create vector of clean variable names
clean_var_names <- c(
  Water_temperature_C = "Water Temperature (°C)",
  Salinity_PSU = "Salinity (PSU)",
  Oxygen_concentration_µmol_L = "Dissolved Oxygen Concentration (µmol/L)",
  RASTERDEPTH_m = "Bathymetric Depth (m)",
  RASTERSLOPE_deg = "Bathymetric Slope (°)",
  NEAR_DIST = "Distance to Nearest Shoreline (m)"
  )
```

```{r, include=FALSE}
# ❌❌❌❌❌❌❌❌❌❌❌❌❌❌❌❌
# Only variables that exist in background
env_vars_bkg <- c(
  "Water_temperature_C",
  "Salinity_PSU",
  "Oxygen_concentration_µmol_L"
  )
```

```{r, include=FALSE}
# ❌❌❌❌❌❌❌❌❌ (env_vars_bkg)
# Pivot longer DETECTION data
hbk_long <- hbk %>%
  select(all_of(env_vars_bkg)) %>%
  pivot_longer(
    cols = everything(),
    names_to = "Variable",
    values_to = "Value"
  ) %>%
  mutate(
    Variable = factor(
      Variable,
      levels = c(
        "Water_temperature_C",
        "Salinity_PSU",
        "Oxygen_concentration_µmol_L"
        )
      )
    ) %>%
  filter(!is.na(Value)) # drop non-finite values
```

```{r, include=FALSE}
# ❌❌❌❌❌❌❌❌❌ (env_vars_bkg)
# # Pivot longer BACKGROUND data
bkg_long <- bkg %>%
  select(all_of(env_vars_bkg)) %>%
  pivot_longer(
    cols = everything(),
    names_to = "Variable",
    values_to = "Value"
  ) %>%
  mutate(
    Variable = factor(
      Variable,
      levels = levels(hbk_long$Variable)  # same order as detections
    )
  ) %>%
  filter(!is.na(Value))  # drop non-finite values
```

```{r, echo=FALSE}
# Explicitly define order of boxes in each plot (detections on left, background on right)
hbk_long$boxorder <- factor("Detection", levels = c("Detection", "Background"))
bkg_long$boxorder <- factor("Background", levels = c("Detection", "Background"))

# Facted box plot
boxplot_COMBINED <- ggplot() +
# DETECTION box (colors by variable, keep outliers)
  geom_boxplot(data = hbk_long, aes(x = boxorder, y = Value, fill = Variable, color = Variable),
               width = 0.9,
               alpha = 0.45,
               outlier.alpha = 0.7,
               outlier.size = 1.5,
               position = position_nudge(x = -0.01)  # shift box slightly left
 ) +
  # BACKGROUND box (all grey, remove outliers)
  geom_boxplot(data = bkg_long, aes(x = boxorder, y = Value),
               fill = "grey80",
               color = "grey40",
               width = 0.9,
               alpha = 0.45,
               outlier.alpha = 0.7,
               outlier.size = 1.5,
               position = position_nudge(x = 0.01)  # shift box slightly right
 ) + 
  facet_wrap(
    ~ Variable,
    scales = "free_y",
    ncol = 3,
    labeller = labeller(Variable = function(x) 
      stringr::str_wrap(clean_var_names[x], width = 18))
 ) + 
  scale_fill_manual(values = custom_colors, guide = "none") +
  scale_color_manual(values = custom_colors, guide = "none") +
  labs(
    x = NULL,
    y = "Value"
  ) +
  theme_minimal(base_size = 10) +
  theme(
    text = element_text(color = "black"),
    strip.text = element_text(color = "black", face = "bold", size = 11),
    axis.text = element_text(color = "black"),
    axis.title = element_text(color = "black"),
    axis.title.y = element_text(face = "bold", size = 10, margin = margin(r = 5)),  # bold y-axis title; r = 5 means y axis label 5 pts away from axis value labels
    axis.text.x = element_text(face = "plain"),  # unbold x-axis labels
    axis.ticks.x = element_blank(),  # remove the tick marks on the x-axis
  )
boxplot_COMBINED
```

**Fig. 2.** \*WRITE CAPTION

```{r, echo=FALSE}
# Faceted desnity plot
density_COMBINED <- ggplot() +
  # BACKGROUND density (grey)
  geom_density(data = bkg_long, aes(x = Value, y = after_stat(density)),
    fill = "grey70",
    color = "grey65",
    alpha = 0.5,
    linewidth = 0.65
  ) +
  # DETECTION density (colored by variable)
  geom_density(data = hbk_long, aes(x = Value, y = after_stat(density), fill = Variable, color = Variable),
    alpha = 0.5,
    linewidth = 0.75
  ) +
  facet_wrap(
    ~ Variable,
    scales = "free",
    ncol = 3,
    labeller = labeller(Variable = function(x) stringr::str_wrap(clean_var_names[x], width = 18))
  ) +
  scale_fill_manual(values = custom_colors, guide = "none") +
  scale_color_manual(values = custom_colors, guide = "none") +
  labs(
    x = "Value",
    y = "Density"
  ) +
  theme_minimal(base_size = 10) +
  theme(
    text        = element_text(color = "black"),
    strip.text  = element_text(color = "black", face = "bold", size = 11),
    axis.text   = element_text(color = "black"),
    axis.title  = element_text(color = "black"),
    axis.title.x = element_text(face = "bold", size = 10, margin = margin(t = 5)),
    axis.title.y = element_text(face = "bold", size = 10, margin = margin(r = 5)),
    axis.ticks.x = element_blank()
  )
density_COMBINED
```

**Fig. 3.** \*WRITE CAPTION
