---
title: "Humpback whale distributions and insight into preferred oceanographic conditions in the Caribbean using passive acoustic monitoring (PAM) and an autonomous underwater vehicle (AUV)"
author: "Keiley Gregory"
date: today
format:
  html:
    css: style.css
    theme: flatly
    highlight-style: tango
execute:
  working-directory: none
bibliography: references.bib
---

***COMMAND F \* FOR FIXES***

Humpback whale (*Megaptera novaeangliae*)

short background on:

how audio recorded by DMON2 and glider mission path

acoustic analysis in LFDCS and manual verification

raster calculations in arc pro

All quantitative oceanographic variables, including those measured directly by sensors on the glider (water temperature, salinity, dissolved oxygen concentration) and those calculated in ArcGIS Pro (bathymetric depth, bathymetric slope, distance to nearest shoreline), ...

future research - study will use data from more missions in different areas, compare findings between missions, overall patterns, blah blah

# Humpback Whale Distributions

short overview of findings from map (where detections clustered, geomorph type, etc.)

**Fig. 1.** \*WRITE CAPTION

\*link to interactive map

# Oceanographic Conditions

overview of findings for preferred oceanographic conditions

```{r, include=FALSE}
library(tidyverse)
library(ggridges) # for ridge plots
library(patchwork) # optional, for combining plots
library(knitr)
library(kableExtra) # optional but makes it prettier
```

```{r, include=FALSE}
# Import humpback detection data with GIS calculations
hbk <- read_csv("data/GISCALC_HUWH_AroundSTX2025.csv")

# Import background mission data
bkg <- read_csv("data/erddap_UVI2_Around_STX_20250306.csv")
```

```{r, include=FALSE}
# Vector of environmental variables to summarize for table (*EXCLUDE glider depth because not really an environmental variable)
env_vars <- c(
  "Water_temperature_C",
  "Salinity_PSU",
  "Oxygen_concentration_µmol_L",
  "RASTERDEPTH_m",
  "RASTERSLOPE_deg",
  "NEAR_DIST"
  )
```

```{r, include=FALSE}
# Create DETECTIONS summary table dataframe
env_summary <- hbk %>%
  select(all_of(env_vars)) %>%
  pivot_longer(
    cols = everything(),
    names_to  = "Variable",
    values_to = "Value"
  ) %>%
  group_by(Variable) %>%
  summarise(
    n     = sum(!is.na(Value)),
    Mean  = mean(Value, na.rm = TRUE),
    SEM   = sd(Value, na.rm = TRUE) / sqrt(n),
    SD    = sd(Value, na.rm = TRUE),
    Min   = min(Value, na.rm = TRUE),
    Max   = max(Value, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  mutate(
    Variable = case_when(
      Variable == "Water_temperature_C" ~ "Water Temperature (°C)",
      Variable == "Salinity_PSU" ~ "Salinity (PSU)",
      Variable == "Oxygen_concentration_µmol_L" ~ "Dissolved Oxygen Concentration (µmol/L)",
      Variable == "RASTERDEPTH_m" ~ "Bathymetric Depth (m)",
      Variable == "RASTERSLOPE_deg" ~ "Bathymetric Slope (°)",
      Variable == "NEAR_DIST" ~ "Distance to Nearest Shoreline (m)",
      TRUE ~ Variable
    )
  )

# Reorder rows for table
env_summary <- env_summary %>%
  mutate(
    Variable = factor(Variable,
    levels = c(
        "Water Temperature (°C)",
        "Salinity (PSU)",
        "Dissolved Oxygen Concentration (µmol/L)",
        "Bathymetric Depth (m)",
        "Bathymetric Slope (°)",
        "Distance to Nearest Shoreline (m)"
      )
    )
  ) %>%
  arrange(Variable)
```

```{r, include=FALSE}
# Create BACKGROUND summary table dataframe
bkg_summary <- bkg %>%
  select("Water_temperature_C", "Salinity_PSU", "Oxygen_concentration_µmol_L") %>%
  pivot_longer(
    cols = everything(),
    names_to  = "Variable",
    values_to = "Value"
    ) %>%
  group_by(Variable) %>%
  summarise(
    n     = sum(!is.na(Value)),
    Mean  = mean(Value, na.rm = TRUE),
    SEM   = sd(Value, na.rm = TRUE) / sqrt(n),
    SD    = sd(Value, na.rm = TRUE),
    Min   = min(Value, na.rm = TRUE),
    Max   = max(Value, na.rm = TRUE),
    .groups = "drop"
    ) %>%
  mutate(
    Variable = case_when(
      Variable == "Water_temperature_C" ~ "Water Temperature (°C)",
      Variable == "Salinity_PSU" ~ "Salinity (PSU)",
      Variable == "Oxygen_concentration_µmol_L" ~ "Dissolved Oxygen Concentration (µmol/L)",
      TRUE ~ Variable
      )
    )

# Reorder rows for table
bkg_summary <- bkg_summary %>%
  mutate(
    Variable = factor(Variable,
    levels = c(
      "Water Temperature (°C)",
      "Salinity (PSU)",
      "Dissolved Oxygen Concentration (µmol/L)"
      )
    )
  ) %>%
  arrange(Variable)
```
